Supplementary methods

### 2020-08-25 text/manuscript.Rmd

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles, and the UMAP which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
the fewest point mutations and culminates in a cluster (class 10) which has
nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
don't observe an obvious trend in the overall mutation burden across the
backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
the molecular-pathway disruption clusters represent advancement through stages,
we projected staging data onto the UMAP. Similar to tissue of origin and other
categories of tumor, we did not observe any bias among the stages to specific
pathway disruption clusters (**figure 3D**), suggesting that stage is not a
contributing factor to cluster identity.

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
pathways in cluster 9, and 0 pathways for clusters 1 and 10. Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

>>>>

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles, and the UMAP which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
the fewest point mutations and culminates in a cluster (class 10) which has
nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
don't observe an obvious trend in the overall mutation burden across the
backbone of the UMAP. Nonetheless, to test the hypothesis that
the molecular-pathway disruption clusters represent advancement through stages,
we projected staging data onto the UMAP. Similar to tissue of origin and other
categories of tumor, we did not observe any bias among the stages to specific
pathway disruption clusters (**figure 3D**), suggesting that stage is not a
contributing factor to cluster identity.

@@

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified 25 pathways highly enriched
(enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2, 9 pathways
enriched in cluster 3, 24 pathways in cluster 4, 10 pathways in cluster 5, 10
pathways in cluster 6, 24 pathways in cluster 7, 24 pathways in cluster 8, 10
pathways in cluster 9, and 0 pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

### 2020-08-25 text/supplemental_methods.Rmd
add page break after SM1 in supp methods

**Figure SM1: Method for tumor classification using pathway aberration data** 
a) MCA analysis was performed to determine the eigenvalue with the most
explanatory power from the categorical data. Average of 100 permutations of the
data was used as a baseline. b) The p-value of each eigenvalue was plotted and
the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
projected down to 3 dimensions for the purpose of density based clustering (only
the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
used to perform hierarchical density based clustering tuned to minimize the
fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
the UMAP projection; tumors are colored by cluster membership (see legend). This
results in some tumors remaining unclassified (grey points). f) kNN was used to
assign membership to the remaining tumors.

@@

Using this model, we split the data by Male/Female sex, and by White, _vs._
Black race as these were the main well-identified categories available to us in
the dataset. Sex is known to be a risk factor for death, with males dying at
slightly higher rates for all ages of interest in our dataset. Black _vs._ White
race is an important category due to differences in access to care and other
socioeconomic and education related factors. However, due to the low number of
Black patient tumors, we use ethnicity only to identify the baseline risk
function, while the tissue- and cluster-specific multipliers are not ethnicity
specific. There were too few tumors from other races to obtain good inference so
we excluded them from our analysis.

>>>>

**Figure SM1: Method for tumor classification using pathway aberration data** 
A) MCA analysis was performed to determine the eigenvalue with the most
explanatory power from the categorical data. Average of 100 permutations of the
data was used as a baseline. B) The p-value of each eigenvalue was plotted and
the max eigenvalue with a pvalue < 0.05 was selected. C) The resulting MCA was
projected down to 3 dimensions for the purpose of density based clustering (only
the first 2 dimensions are shown). Each dot represents a tumor D) HDBSCAN was
used to perform hierarchical density based clustering tuned to minimize the
fraction of unassigned tumors. E) HDBSCAN clustered tumors were visualized on
the UMAP projection; tumors are colored by cluster membership (see legend). This
results in some tumors remaining unclassified (grey points). F) kNN was used to
assign membership to the remaining tumors.

@@

Using this model, we split the data by Male/Female sex, and by White _vs._
Black race as these were the main well-identified categories available to us in
the dataset. Sex is known to be a risk factor for death, with males dying at
slightly higher rates for all ages of interest in our dataset. Black _vs._ White
race is an important category due to differences in access to care and other
socioeconomic and education related factors. However, due to the low number of
Black patient tumors, we use ethnicity only to identify the baseline risk
function, while the tissue- and cluster-specific multipliers are not ethnicity
specific. There were too few tumors from other races to obtain good inference so
we excluded them from our analysis.
